# Reproducible Research: Peer Assessment 1
===========================================
## Overview
This report describes the steps taken to demonstrate Reproducible Research for the Coursera Reproducible
Research Peer Assessment 1. It makes use of data from a personal activity monitoring device. The device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day. 

By analyzing this data, the report answers the questions as to the average number of steps taken each day, the daily activity patterns and highlights the differences between weekday and weekend activity patterns.

Additionally the report shows the steps taken and R Code used to provide the analysis results.

## Loading and preprocessing the data
1) Download and extract the [activity.zip](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) file into the R working directory.

2) Read the data into R as a data frame.

```{r readdate, echo = TRUE}
        ## Load the data
        df <- read.csv("activity.csv", header = TRUE)
```

3) View the raw data.

```{r viewdata, echo = TRUE,}

## Summarize the raw data
summary(df)

## View the data stucture
str(df)

```

4) Transform the data for analysis.

```{r transformdata, echo = TRUE}

## Transform the column data is of a usable format, i.e steps column is numeric,
## date colum is date format and the interval colum is numeric.
df$steps <- as.numeric(df$steps)
df$date <- as.character(df$date)
df$interval <- as.numeric(df$interval)

## Convert the date colum to the appropriately formatted date
df$date <- as.Date(df$date, format = "%Y-%m-%d")

```


## What is mean total number of steps taken per day?
To analyse the number of steps taken per day, the data is broken down by day and totaled. The mean and median can then be calculated.

```{r sumbydate, echo = TRUE}
## Summarize the number of steps by day
sumByDate <- tapply(df$steps, df$date, sum)

## Get the Mean of the number of steps per day
meanByDate <- mean(sumByDate, na.rm = TRUE)

## Get the Median of the numbner of steps per day
medianByDate <- median(sumByDate, na.rm = TRUE)

```

Below is the histogram depicting the mean total number of steps taken per day.

```{r histogram1, echo = TRUE}
## Plot the histogram showing the number steps taken per day.
hist(sumByDate, xlab = "No. steps taken per day", col = "red",
     main = "Histogram depicting the number of steps taken per day")

```

The Mean is: **`r format(meanByDate, scientific = F)`**  
The Median is: **`r format(medianByDate, scientific = F)`**

## What is the average daily activity pattern?
To analyse the average daily activity pattern, the number of steps per daily interval are averaged. Below is a Time-Series plot depicting this pattern.

```{r activityply, echo = TRUE}
## Summarize the number of steps per interval and the mean
sumByInterval <- tapply(df$steps, df$interval, sum)
meanByInterval <- tapply(df$steps, df$interval, mean, na.rm = TRUE)

## Plot the pattern of daily activity
plot(unique(df$interval), meanByInterval, type = "l", col = "red", 
     xlab = "5-minute intervals", ylab = "Average no. steps taken for all days")

## Find the interval containing the maximum number of steps on average
interval <- df[which.max(meanByInterval), 3]

```

On average across all the days, the 5-minute interval that contains the maximum amount of steps, is the **`r interval`th** interval.

## Imputing missing values
From the initial summary taken it is evident that there are **2304** missing values (See item **3.** of the **Loading and processing the data** section). This can be verified from the following code chunk:

```{r missing, echo = TRUE}
## Confirm the total number of missing values (NA) in the data set
totalMissing <- sum(is.na(df))

```
Total number of missing values: **`r totalMissing`**

These missing values introduce bias into the above 







since we know that the average number of steps (with missing values removed), is **`r format(meanByInterval, scientific = F)`**, we can create a new data frame and substitutre any missing values with the mean value. 





## Are there differences in activity patterns between weekdays and weekends?
