# Reproducible Research: Peer Assessment 1
===========================================
## Overview
This report describes the steps taken to demonstrate Reproducible Research for the Coursera Reproducible
Research Peer Assessment 1. It makes use of data from a personal activity monitoring device. The device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day. 

By analyzing this data, the report answers the questions as to the average number of steps taken each day, the daily activity patterns and highlights the differences between weekday and weekend activity patterns.

Additionally the report shows the steps taken and R Code used to provide the analysis results.

## Loading and preprocessing the data
1) Download and extract the [activity.zip](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) file into the R working direcory.

2) Read the data into R as a data frame.

```{r readdate, echo = TRUE}
        ## Load the data
        df <- read.csv("activity.csv", header = TRUE)
```

3) View the raw data.

```{r viewdata, echo = TRUE,}

## Summarize the raw data
summary(df)

## View the data stucture
str(df)

```

4) Transform the data for analysis.

```{r transformdata, echo = TRUE}

## Transform the column data is of a usable format, i.e steps column is numeric,
## date colum is date format and the interval colum is numeric.
df$steps <- as.numeric(df$steps)
df$date <- as.character(df$date)
df$interval <- as.numeric(df$interval)

## Convert the date colum to the appropriately formatted date
df$date <- as.Date(df$date, format = "%Y-%m-%d")

```


## What is mean total number of steps taken per day?
To analyse the number of steps taken per day, the data is broken down and summarized by.

```{r sumbydate, echo = TRUE}
## Summarize the number of steps by day
sumByDate <- tapply(df$steps, df$date, sum)

## Get the Mean of the number of steps per day
meanByDate <- mean(sumByDate, na.rm = TRUE)

## Get the Median of the numbner of steps per day
medianByDate <- median(sumByDate, na.rm = TRUE)

```

Below is the histogram detailing the mean total number of steps taken per day.

```{r histogram1, echo = TRUE}
## Plot the histogram showing the number steps taken per day.
hist(sumByDate, xlab = "No. steps taken per day", col = "red",
     main = "Histogram showing the number of steps taken per day")

```


## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
